---
title: "Installation Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<br></br>

# Installation

---

The `metapsyTools` package has not been submitted to [CRAN](https://cran.r-project.org/),
which means that an installation using the `install.packages()` function will not 
work. 

To install the package, you have to download the latest version of the package from the 
`metapsyTools` [Github repository](https://github.com/MathiasHarrer/metapsyTools). This is fairly
easy to achieve using the `install_github` function in the 
[`devtools`](https://devtools.r-lib.org/index.html) package. If you run the following code in R:

```{r, eval=F}
if (!require("devtools"))
  install.packages("devtools")

devtools::install_github(
    "MathiasHarrer/metapsyTools",
    build_vignettes = TRUE)
```

R will automatically check if `devtools` is installed on your computer, and install it
for you if necessary; download the `metapsyTools` package from Github; and install the package. 
Following a succesfull installation, the package can be loaded from your library and is then 
ready to be used.

```{r, eval=F}
library(metapsyTools)
```

During the installation process, you may receive the following **prompts**:

```
## These packages have more recent versions available.
## Which would you like to update?
## 
## 1: All                          
## 2: CRAN packages only            
## 3: None                          
## 4: ggpubr (0.2.2 -> 0.2.3) [CRAN]
## 5: zip    (2.0.3 -> 2.0.4) [CRAN]
## 
## Enter one or more numbers, or an empty line to skip updates:
```

When you get this message, it is best to tell the installation manager that no packages should be updated. 
In our example, this means pasting `3` into the console and then hitting Enter. 
In the same vein, when the installation manager asks this question:

```
## There are binary versions available but the source versions are later:
##  
##  [...]
##  
##   Do you want to install from sources the package which needs compilation?
##   y/n: 
```

It is best to choose `n` (no). If the installation fails with this strategy 
(meaning that you get an Error), run the installation again, but update all packages this time.


<br></br>

# Updating the Package

---

The `metapsyTools` package is under active development and after some time, a new 
version may be available. You can check the current version of of `metapsyTools`
installated on your computer using this code:

```{r, eval=F}
packageVersion("metapsyTools")
```

The header of this website displays the most recent version of the package available
for users. To use novel functionalities integrated into the package since your last installation,
you will have to update the package that you have locally on your computer. 

To update your package version, you first have to de-install the current version on your computer.
You can do this using this code:

```{r, eval=F}
# Note: this code assumes you are working with RStudio 
remove.packages("metapsyTools")
.rs.restartR() 
```

You can then use the same code as used above to (re-)install the latest version
of the package.

<br></br>